---
title: "VISTA Practical Workshop"
subtitle: "Hands-on RNA-seq workflow using VISTA example data"
author: "VISTA workshop"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## Learning Objectives

By the end of this practical you should be able to:

- Load VISTA and its example RNA-seq data.
- Inspect count data and sample metadata.
- Prepare counts and metadata for VISTA.
- Create and validate a VISTA object.
- Understand where normalized expression, sample metadata, gene metadata, and DE results are stored.
- Make QC, differential expression, and expression plots.
- Run functional enrichment after checking DE results.
- Export tables and plots.

## Setup

Run this chunk first.

```{r setup}
library(VISTA)
library(ggplot2)
library(dplyr)
library(tibble)
```

Some later sections use optional Bioconductor packages. This helper lets us skip a section cleanly if a package is not installed.

```{r helper-functions}
has_pkg <- function(pkg) {
  requireNamespace(pkg, quietly = TRUE)
}
```

## 1. Load the Example Data

VISTA includes two example datasets:

- `count_data`: RNA-seq counts.
- `sample_metadata`: sample-level information.

```{r load-example-data}
data(count_data, package = "VISTA")
data(sample_metadata, package = "VISTA")
```

Inspect the size of each object.

```{r inspect-dimensions}
dim(count_data)
dim(sample_metadata)
```

Preview the first few rows.

```{r preview-data}
head(count_data[, 1:5])
head(sample_metadata)
```

## 2. Understand the Inputs

The count table has one row per gene and one column per sample, plus a `gene_id` column.

```{r count-columns}
colnames(count_data)
```

The metadata has one row per sample. The `sample_names` column should match the count columns.

```{r metadata-columns}
names(sample_metadata)
sample_metadata$sample_names
```

Check the treatment column.

```{r treatment-summary}
table(sample_metadata$dex)
```

For this workshop, we compare:

- numerator: `trt`
- denominator: `untrt`
- group column: `dex`

In plain language: treated samples compared with untreated samples.

## 3. Prepare Counts and Metadata

VISTA provides helper functions to standardize inputs before analysis.

```{r prepare-counts}
prepared_counts <- read_vista_counts(
  count_data,
  format = "matrix",
  gene_id_column = "gene_id"
)
```

```{r prepare-metadata}
prepared_metadata <- read_vista_metadata(
  sample_metadata,
  sample_column = "sample_names"
)
```

Now match counts and metadata. This is one of the most important checks in RNA-seq analysis.

```{r match-inputs}
matched_inputs <- match_vista_inputs(
  counts = prepared_counts,
  sample_info = prepared_metadata,
  column_geneid = "gene_id",
  sample_column = "sample_names"
)
```

Inspect the matched output.

```{r inspect-matched-inputs}
names(matched_inputs)
dim(matched_inputs$counts)
dim(matched_inputs$sample_info)
matched_inputs$sample_info[, c("sample_names", "dex", "cell")]
```

## 4. Create a VISTA Object

Now we run differential expression and create a VISTA object.

For beginners, focus on these arguments:

- `counts`: the gene count table.
- `sample_info`: the sample metadata.
- `column_geneid`: the gene ID column.
- `group_column`: the metadata column defining groups.
- `group_numerator`: the group in the numerator of the fold change.
- `group_denominator`: the reference group.
- `method`: the differential expression backend.

```{r create-vista}
vista <- create_vista(
  counts = matched_inputs$counts,
  sample_info = matched_inputs$sample_info,
  column_geneid = matched_inputs$column_geneid,
  group_column = "dex",
  group_numerator = "trt",
  group_denominator = "untrt",
  method = "deseq2",
  min_counts = 10,
  min_replicates = 2,
  log2fc_cutoff = 1,
  pval_cutoff = 0.05,
  p_value_type = "padj"
)
```

Print the object.

```{r print-vista}
vista
```

## 5. Validate the Object

Validation checks whether the VISTA object is internally consistent.

```{r validate-vista}
validate_vista(vista, level = "full")
```

## 6. Why the VISTA Object Matters

A VISTA object keeps several parts of the analysis together:

- normalized expression values,
- sample metadata,
- gene metadata,
- differential expression results,
- analysis cutoffs,
- color settings and plotting information.

This follows the same logic as Bioconductor's `SummarizedExperiment`: data and metadata should stay synchronized.

Inspect key parts:

```{r inspect-vista-parts}
SummarizedExperiment::colData(vista) |> as.data.frame() |> head()
SummarizedExperiment::rowData(vista) |> as.data.frame() |> head()
names(comparisons(vista))
deg_summary(vista)
cutoffs(vista)
```

## 7. Add Gene Annotations

The example data uses Ensembl gene IDs. For human-readable plots, gene symbols are useful.

This step uses `org.Hs.eg.db`. If the package is not installed, skip this section and continue with Ensembl IDs.

```{r annotate-genes}
if (has_pkg("org.Hs.eg.db")) {
  vista <- set_rowdata(
    vista,
    orgdb = org.Hs.eg.db::org.Hs.eg.db,
    columns = c("SYMBOL", "GENENAME", "ENTREZID"),
    keytype = "ENSEMBL"
  )
}

SummarizedExperiment::rowData(vista) |> as.data.frame() |> head()
```

## 8. Explore Differential Expression Results

Get the comparison name.

```{r comparison-name}
comp_name <- names(comparisons(vista))[1]
comp_name
```

Extract the DE result table.

```{r de-results}
de_table <- comparisons(vista)[[comp_name]]
head(de_table)
```

Count genes by regulation category.

```{r regulation-counts}
table(de_table$regulation)
```

Get the top upregulated and downregulated genes.

```{r top-genes}
top_up <- get_genes_by_regulation(
  vista,
  sample_comparisons = comp_name,
  regulation = "Up",
  top_n = 20
)[[1]]

top_down <- get_genes_by_regulation(
  vista,
  sample_comparisons = comp_name,
  regulation = "Down",
  top_n = 20
)[[1]]

top_up
top_down
```

## 9. Quality Control Plots

QC plots help us check whether samples behave as expected.

### Sample Correlation Heatmap

```{r corr-heatmap, fig.width=7, fig.height=6}
get_corr_heatmap(
  vista,
  viridis_direction = -1,
  show_corr_values = TRUE,
  label_size = 3
)
```

### PCA Plot

```{r pca-plot, fig.width=7, fig.height=5}
get_pca_plot(
  vista,
  label = TRUE,
  shape_by = "cell"
)
```

### MDS Plot

```{r mds-plot, fig.width=7, fig.height=5}
get_mds_plot(
  vista,
  label = TRUE,
  shape_by = "cell"
)
```

### UMAP Plot

UMAP is optional and requires the `uwot` package.

```{r umap-plot, fig.width=7, fig.height=5}
if (has_pkg("uwot")) {
  get_umap_plot(
    vista,
    label = TRUE,
    shape_by = "cell"
  )
}
```

## 10. Differential Expression Plots

### DEG Count Barplot

```{r deg-count-barplot, fig.width=7, fig.height=5}
get_deg_count_barplot(vista)
```

### Volcano Plot

```{r volcano-plot, fig.width=8, fig.height=7}
if (has_pkg("EnhancedVolcano")) {
  get_volcano_plot(
    vista,
    sample_comparison = comp_name,
    display_id = "SYMBOL"
  )
}
```

### MA Plot

```{r ma-plot, fig.width=8, fig.height=6}
get_ma_plot(
  vista,
  sample_comparison = comp_name,
  label_n = 10,
  display_id = "SYMBOL"
)
```

## 11. Expression Plots for Selected Genes

Expression plots help us inspect individual genes or sets of genes.

```{r selected-genes}
selected_genes <- head(top_up, 4)
selected_genes
```

### Expression Heatmap

```{r expression-heatmap, fig.width=8, fig.height=8}
if (has_pkg("ComplexHeatmap")) {
  get_expression_heatmap(
    vista,
    genes = c(head(top_up, 10), head(top_down, 10)),
    sample_group = c("untrt", "trt"),
    display_id = "SYMBOL",
    show_row_names = TRUE,
    summarise_replicates = FALSE,
    annotate_columns = c("dex", "cell")
  )
}
```

### Barplot

```{r expression-barplot, fig.width=9, fig.height=6}
get_expression_barplot(
  vista,
  genes = selected_genes,
  display_id = "SYMBOL",
  log_transform = TRUE,
  stats_group = TRUE
) +
  theme_minimal(base_size = 14)
```

### Boxplot

```{r expression-boxplot, fig.width=9, fig.height=6}
get_expression_boxplot(
  vista,
  genes = selected_genes,
  display_id = "SYMBOL",
  log_transform = TRUE,
  facet_by = "gene",
  stats_group = TRUE
) +
  theme_minimal(base_size = 14)
```

### Violin Plot

```{r expression-violin, fig.width=9, fig.height=6}
get_expression_violinplot(
  vista,
  genes = selected_genes,
  display_id = "SYMBOL",
  log_transform = TRUE,
  facet_by = "gene"
) +
  theme_minimal(base_size = 14)
```

## 12. Fold-Change Analysis

Extract a matrix of log2 fold changes.

```{r foldchange-matrix}
fc_matrix <- get_foldchange_matrix(vista)
head(fc_matrix)
```

Plot fold changes for selected genes.

```{r foldchange-heatmap, fig.width=7, fig.height=8}
if (has_pkg("ComplexHeatmap")) {
  get_foldchange_heatmap(
    vista,
    genes = c(head(top_up, 10), head(top_down, 10)),
    display_id = "SYMBOL",
    show_row_names = TRUE
  )
}
```

## 13. Before Functional Analysis

Functional analysis should be run after checking the DE result table.

Before running enrichment, confirm:

- Which comparison is being used.
- Which genes are Up, Down, or Other.
- Which cutoffs define significant genes.
- Which gene ID type is available.
- Whether annotation packages are installed.

```{r before-functional}
comp_name
cutoffs(vista)
table(de_table$regulation)
head(SummarizedExperiment::rowData(vista) |> as.data.frame())
```

For this dataset, our original gene IDs are Ensembl IDs. We therefore use `from_type = "ENSEMBL"` in enrichment functions.

## 14. MSigDB Enrichment

Run Hallmark enrichment for upregulated genes.

```{r msigdb-enrichment}
if (has_pkg("msigdbr") && has_pkg("clusterProfiler") && has_pkg("org.Hs.eg.db")) {
  msig_up <- get_msigdb_enrichment(
    vista,
    sample_comparison = comp_name,
    regulation = "Up",
    msigdb_category = "H",
    species = "Homo sapiens",
    from_type = "ENSEMBL",
    orgdb = org.Hs.eg.db::org.Hs.eg.db
  )

  head(msig_up$enrich@result)
}
```

Plot the enrichment result.

```{r msigdb-plot, fig.width=9, fig.height=6}
if (exists("msig_up") && !is.null(msig_up$enrich) && nrow(msig_up$enrich@result) > 0) {
  get_enrichment_plot(msig_up$enrich, top_n = 10)
}
```

## 15. GO Enrichment

Run Gene Ontology Biological Process enrichment.

```{r go-enrichment}
if (has_pkg("clusterProfiler") && has_pkg("org.Hs.eg.db")) {
  go_bp <- get_go_enrichment(
    vista,
    sample_comparison = comp_name,
    regulation = "Up",
    ont = "BP",
    species = "Homo sapiens",
    from_type = "ENSEMBL",
    orgdb = org.Hs.eg.db::org.Hs.eg.db
  )

  head(go_bp$enrich@result)
}
```

```{r go-plot, fig.width=9, fig.height=6}
if (exists("go_bp") && !is.null(go_bp$enrich) && nrow(go_bp$enrich@result) > 0) {
  get_enrichment_plot(go_bp$enrich, top_n = 10)
}
```

## 16. GSEA

GSEA uses a ranked list of genes instead of only a significant-gene list.

```{r gsea}
if (has_pkg("msigdbr") && has_pkg("clusterProfiler") && has_pkg("org.Hs.eg.db")) {
  gsea_h <- get_gsea(
    vista,
    sample_comparison = comp_name,
    set_type = "msigdb",
    from_type = "ENSEMBL",
    species = "Homo sapiens",
    msigdb_category = "H",
    orgdb = org.Hs.eg.db::org.Hs.eg.db,
    pvalueCutoff = 0.05
  )

  head(gsea_h$enrich@result)
}
```

```{r gsea-plot, fig.width=9, fig.height=6}
if (exists("gsea_h") && !is.null(gsea_h$enrich) && nrow(gsea_h$enrich@result) > 0) {
  get_enrichment_plot(gsea_h$enrich, top_n = 10)
}
```

## 17. Pathway-Specific Heatmap

Use genes from enriched pathways and visualize their expression.

```{r pathway-heatmap, fig.width=9, fig.height=8}
if (exists("msig_up") &&
    !is.null(msig_up$enrich) &&
    nrow(msig_up$enrich@result) > 0 &&
    has_pkg("ComplexHeatmap")) {

  get_pathway_heatmap(
    x = vista,
    enrichment = msig_up$enrich,
    sample_group = c("untrt", "trt"),
    top_n = 2,
    gene_mode = "union",
    max_genes = 50,
    value_transform = "zscore",
    display_id = "SYMBOL",
    annotate_columns = c("dex", "cell"),
    summarise_replicates = FALSE,
    show_row_names = FALSE
  )
}
```

## 18. Export Results

Create an output directory.

```{r create-output-dir}
dir.create("workshop_outputs", showWarnings = FALSE)
```

Save the main DE result table.

```{r save-de-results}
save_vista_data(
  vista,
  what = "comparison",
  sample_comparison = comp_name,
  file = file.path("workshop_outputs", "vista_de_results.csv")
)
```

Save a plot.

```{r save-plot}
p <- get_ma_plot(
  vista,
  sample_comparison = comp_name,
  label_n = 10,
  display_id = "SYMBOL"
)

save_vista_plot(
  p,
  file = file.path("workshop_outputs", "ma_plot.png"),
  width = 8,
  height = 6
)
```

Export a bundle of standard assets.

```{r export-assets, eval=FALSE}
export_vista_assets(
  vista,
  out_dir = "workshop_outputs/vista_assets",
  sample_comparison = comp_name,
  display_id = "SYMBOL"
)
```

Save the VISTA object for later.

```{r save-vista-object}
saveRDS(vista, file = file.path("workshop_outputs", "airway_vista.rds"))
```

## 19. Exercises

Try these after the main workflow runs. Start with the easy questions, then move to moderate and hard questions if you have time.

### Easy Questions

These questions focus on reading objects, changing simple arguments, and understanding the output.

1. How many genes and samples are present in `count_data`?
2. Which column in `sample_metadata` stores the sample names used by VISTA?
3. How many samples are in each `dex` group?
4. What is the name of the comparison stored in `comp_name`?
5. How many genes are labelled `Up`, `Down`, and `Other` in `de_table$regulation`?
6. Change `label_n` in the MA plot from 10 to 20. What changes in the plot?
7. Make a PCA plot without sample labels by changing `label = TRUE` to `label = FALSE`.
8. Change the point size in the PCA plot. Which value makes the plot easiest to read?
9. Select the top 5 upregulated genes and print them.
10. Select the top 5 downregulated genes and print them.

### Moderate Questions

These questions require modifying code and interpreting the result.

1. Create a barplot for the top 5 downregulated genes using `get_expression_barplot()`.
2. Create a boxplot for the top 5 upregulated genes using `get_expression_boxplot()`.
3. Change the expression heatmap so that it summarizes replicate groups instead of showing every sample.
4. Make the expression heatmap using only the top 10 upregulated genes.
5. Change the MA plot so that no genes are labelled. Which argument controls this?
6. Run MSigDB enrichment for downregulated genes instead of upregulated genes.
7. Make an enrichment plot for the downregulated MSigDB result.
8. Export the DE result table using a new filename, such as `vista_de_results_updated.csv`.
9. Save a PCA plot as a PNG file using `save_vista_plot()`.
10. Compare the PCA and MDS plots. Do the samples group in a similar way?

### Hard Questions

These questions are designed for deeper thinking, troubleshooting, and extending the workflow.

1. Change the differential expression cutoff in `create_vista()` from `log2fc_cutoff = 1` to `log2fc_cutoff = 0.5`. How does the number of Up and Down genes change?
2. Change `pval_cutoff = 0.05` to `pval_cutoff = 0.01`. How does this affect the DEG summary?
3. Re-run the workflow using `method = "edger"` instead of `method = "deseq2"`. Are the top genes similar?
4. Add `cell` as a covariate in `create_vista()` using `covariates = c("cell")`. How does this change the results?
5. Try `method = "both"` and inspect the available DE result sources. What is the difference between using `deseq2`, `edger`, and `consensus` results?
6. Use `set_de_source()` to switch the active DE source if you created a multi-method VISTA object.
7. Run GO enrichment for downregulated genes and compare the top terms with the upregulated GO results.
8. Use `get_pathway_genes()` to extract genes from the top enriched pathway. Then make an expression heatmap for those genes.
9. Create a small report of your top biological interpretation: one sentence about the comparison, one sentence about the strongest DE genes, and one sentence about the top enriched pathway.
10. Think about reproducibility: which files would you save or share so someone else can repeat your analysis?

## 20. Key Takeaways

- Always inspect data before analysis.
- Sample names in counts and metadata must match.
- A VISTA object keeps expression, metadata, annotations, and results together.
- QC plots should be checked before interpretation.
- Functional analysis should be run after DE results and gene IDs are understood.
- Exported tables and plots make the analysis easier to share and reproduce.

## 21. Additional Resources

- [VISTA Bioconductor package page](https://www.bioconductor.org/packages//release/bioc/html/VISTA.html)
- [VISTA pkgdown documentation](https://cparsania.github.io/VISTA/)
- [VISTA GitHub repository](https://github.com/cparsania/VISTA/)

## Session Information

```{r session-info}
sessionInfo()
```
